---
title: "Bellabeat Case Study"
output: html_document
date: "2025-12-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readr)
```

Introduction (ASK Phase)
Bellabeat is a high‑tech wellness company that manufactures smart devices focused on women's health.
The goal of this case study is to analyze smart‑device usage data to identify trends in user activity, sleep, and daily habits.
These insights will help Bellabeat strengthen its marketing strategy and support the growth of the Leaf wellness tracker.



#Prepare phase

```{r  load_date, include=FALSE}
daily_activity <- read_csv("dailyActivity_merged.csv")
sleep_day <- read_csv("sleepDay_merged.csv")

str(daily_activity)
str(sleep_day)
```

#process phase

Convert date columns from character to Date format


```{r}
daily_activity$ActivityDate <- as.Date(daily_activity$ActivityDate, format = "%m/%d/%Y")
sleep_day$SleepDay <- as.Date(sleep_day$SleepDay, format = "%m/%d/%Y %I:%M:%S %p")

str(daily_activity)
str(sleep_day)
```

🔍 Exploratory Data Analysis (EDA)

View column names and first row

```{r}
head(daily_activity)
colnames(daily_activity)

head(sleep_day)
colnames(sleep_day)
```

Number of unique users

```{r}
n_distinct(daily_activity$Id)
n_distinct(sleep_day$Id)
```

Number of observations

```{r}
nrow(daily_activity)
nrow(sleep_day)
```

Summary Statistics

Daily Activity Summary

```{r}

daily_activity %>%
  select(TotalSteps, TotalDistance, SedentaryMinutes) %>%
  summary()

```

Sleep Summary

```{r}
sleep_day %>%
  select(TotalSleepRecords, TotalMinutesAsleep, TotalTimeInBed) %>%
  summary()

```


📈 Visual Explorations (ANALYZE Phase)

Steps vs calories

```{r}
ggplot(daily_activity, aes(TotalSteps, Calories)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  labs(title = "Steps vs Calories") +
  theme_minimal()
```

Users who take more steps tend to burn more calories.
This confirms that physical activity is the main driver of energy expenditure

Minutes Asleep vs Time in Bed

```{r}
ggplot(sleep_day, aes(TotalMinutesAsleep, TotalTimeInBed)) +
  geom_point(alpha = 0.6, color = "blue") +
  labs(title = "Minutes Asleep vs Time in Bed") +
  theme_minimal()
```

There is a strong positive relationship between time in bed and minutes asleep.
However, the scatter shows that users often spend time awake while in bed, indicating varying sleep efficiency




🔗 Merge Datasets(Activity vs Sleep)

```{r}

combined_data <- merge(
  sleep_day,
  daily_activity,
  by.x = c("Id", "SleepDay"),
  by.y = c("Id", "ActivityDate")
)

n_distinct(combined_data$Id)
combined_data

combined_data %>%
  mutate(weekday = weekdays(SleepDay)) %>%
  group_by(weekday) %>%
  summarise(
    avg_steps = mean(TotalSteps, na.rm = TRUE),
    avg_sleep = mean(TotalMinutesAsleep, na.rm = TRUE)
  )


```

Activity and sleep patterns vary by weekday.
This can help Bellabeat tailor notifications or coaching features based on user behavior



```{r}

combined_data <- combined_data %>%
  mutate(sleep_efficiency = TotalMinutesAsleep / TotalTimeInBed)

summary(combined_data$sleep_efficiency)

ggplot(combined_data, aes(sleep_efficiency)) +
  geom_histogram(bins = 30, fill = "purple", alpha = 0.7) +
  labs(title = "Distribution of Sleep Efficiency",
       x = "Sleep Efficiency",
       y = "Count") +
  theme_minimal()

```

Most users have moderate sleep efficiency, but the distribution shows variability.
Some users spend a significant portion of time in bed awake, suggesting opportunities for sleep‑improvement coaching.


Minutes Asleep vs Sedentary Minutes

```{r}

ggplot(data=combined_data, aes(x=TotalMinutesAsleep, y=SedentaryMinutes)) + 
geom_point(color='darkblue') + geom_smooth() +
  labs(title="Minutes Asleep vs. Sedentary Minutes")

```
There is no strong relationship between sedentary time and sleep duration.
Sedentary behavior alone does not predict sleep quality.


Steps vs Minutes Asleep

```{r}
ggplot(combined_data, aes(TotalSteps, TotalMinutesAsleep)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm") +
  theme_minimal()

```

There is a weak positive trend: users who walk more may sleep slightly longer.
However, the relationship is not strong enough to be conclusive.


Histogram of Minutes Asleep
```{r}
ggplot(combined_data, aes(TotalMinutesAsleep)) +
  geom_histogram(bins = 30, fill = "steelblue") +
  theme_minimal()
```

Most users sleep between 350–450 minutes (about 6–7.5 hours).
This is slightly below the recommended 7–9 hours of sleep.


# SHARE Phase: Communicating Insights

The analysis reveals several important trends in user behavior:

- Users are generally not meeting recommended activity levels, averaging only 6,547 steps per day.
- Sedentary behavior is high, with a median of 1,057 sedentary minutes per day.
- Sleep duration is slightly below recommended levels, and sleep efficiency varies significantly.
- Activity and sleep show only weak correlations, suggesting that users may need more structured guidance.
- Weekday patterns indicate opportunities for targeted engagement.

These insights help Bellabeat understand how users interact with their wellness data and where behavior improvements can be encouraged.


# ACT Phase: Recommendations for Bellabeat

Based on the analysis, several strategic actions can help Bellabeat improve user engagement and support healthier habits:

### 1. Promote increased physical activity
Bellabeat can introduce step‑based challenges, personalized reminders, and motivational notifications to help users reach the recommended 10,000 steps per day.

### 2. Support better sleep habits
Since sleep efficiency varies widely, Bellabeat can offer bedtime routine guidance, sleep hygiene tips, and insights when users spend too much time awake in bed.

### 3. Personalize engagement based on weekday patterns
Activity and sleep behaviors differ across the week. Bellabeat can tailor notifications and coaching to match these patterns.

### 4. Highlight holistic wellness features
Marketing should emphasize stress tracking, mindfulness, hydration reminders, and cycle tracking — features that align with user needs.

### 5. Build data‑driven marketing campaigns
Insights from real user behavior can shape campaigns that resonate with women seeking healthier routines.

These actions can help Bellabeat strengthen its brand, improve user engagement, and support long‑term wellness outcomes.


```
























